Implicit yaddayadda

### 2011-10-28 papers/PLDI2012/implicitization.tex
tassage colonnes
Améliorations mineures et remarques

\JH{$99990)$, ca designe bien une borne d'intervalle ouvert ? J'ai plus
l'habitude a la notation $99990[$}
Widening iterations converge to $\lstinline|x_old| \in (-\infty,+\infty)$; let us now see why decreasing iterations fail to recover the desired invariant.
The \lstinline|x > x_old+10| test at line~6, if taken, yields $\lstinline|x_old|
\in (-\infty,99990)$; followed by \lstinline|x = x_old+10|, we obtain
$\lstinline|x| \in (-\infty,\allowbreak 100000)$, and the same after union with the no-operation ``else'' branch. Line~7 yields $\lstinline|x| \in (-\infty,+\infty)$.

@@

For a loop header $p_i$, this technique computes a disjunctive invariant
$\bigvee_{1\leq j \leq m} X_{i,j}$. 
To do so, one chooses an integer $\delta \in [1,m]$, and
a mapping function $\sigma: [1,m] \times [1,n] \mapsto [1,m]$, assuming there
are $n$ distinct paths in the loop.  For each
abstract value of the disjunctive invariant, and for each path $\tau_{i,k}$ in the loop, the
image of $X_{i,j}$ by the path is joined with
$X_{i,\sigma(j,k)}$.
Initially, the $\delta$-th abstract value contains the initial states of
$p_i$, and all other abstract values contain~$\emptyset$.

@@

Our improvement consists in a modification of the SMT formula we solve in
\ref{sec:guided_multigraph}.
We introduce in this formula Boolean variables $\{d_j, 1 \leq j \leq m\}$, so
that we can easily find in the model which abstract value of the disjunction has
to be chosen to make the invariant grow.
The resulting formula that is given to the SMT solver is defined
by $g(p_i)$.
When the formula is satisfiable, we know that the index $j$ of the starting
disjunct that has to be chosen is the one for which the associate Boolean value
$d_j$ is \emph{true} in the model. Then, we can easily compute the value of 
$\sigma(j,k)$, thus know the index of the disjunct to join with.

@@

@@

\subsection{Analysis algorithm}
\label{sec:analysis-algorithm}
LLVM bitcode is in \emph{single static assignment} (SSA) form: a given scalar variable is given a value at a single syntactic point in the program. In concrete terms, an assignment \lstinline|x=2*x+1;| gets translated into a definition $x_2 = 2x_1+1$, with distinct variables $x_1$ and $x_2$ corresponding to the same original variable \lstinline|x| at different points in the program.
LLVM makes it easy to follow definition-use and use-definition chains: for a given variable (say, $x_2$) one can immediately obtain its definition (say, $2x_1+1$).
One may see conversion to SSA form as a static precomputation of some of the symbolic propagations proposed by \citet{DBLP:conf/vmcai/Mine06} to enhance the precision of analyses.

@@

For the computation of disjunctive invariants, we experimented with a heuristic
of dynamic construction of the $\sigma$ function, adapted from
\cite{DBLP:conf/pldi/GulwaniZ10}.  We start with one single
disjunct per control point ($m=1$), and we start with an empty map $\sigma$. We also
define a maximal bound on the number of disjunct per control point, noted $M$.
$\sigma(j,k)$ is then constructed on the fly only when needed.
When $\sigma(j,k)$ needs to be computed, we first compute the image of the
abstract value $X_k$ by the path indexed by $j$, and try to find an existing
disjunct of index $k'$ so that the least upper bound of the two abstract values is
exactly the union of them. If such an index exists, then $\sigma(j,k) = k'$.
Otherwise:
\begin{itemize}
	\item if $m < M$, we increase $m$ by $1$ and define $\sigma(j,k) = m$
	\item if $m = M$, we define $\sigma(j,k) = M$
\end{itemize}

@@

\begin{table*}
\begin{center}
\setlength{\tabcolsep}{1ex}
\begin{tabular}{|l
|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}%
|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}%
|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}%
|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}%
|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}%
|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}|} \hline
\multicolumn{1}{|c|}{\textbf{Benchmark}}
& \multicolumn{3}{c|}{\textbf{G/S}}
& \multicolumn{3}{c|}{\textbf{PF/S}}
& \multicolumn{3}{c|}{\textbf{PF/G}}
& \multicolumn{3}{c|}{\textbf{G+PF/PF}}
& \multicolumn{3}{c|}{\textbf{G+PF/G}}
& \multicolumn{3}{c|}{\textbf{DIS/G+PF}} \\ %\cline{2-16}
& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.} \\
 \hline
 \input{table1}
\end{tabular}
\end{center}
\caption{Result of the comparison of the various techniques described in this
paper: classic Abstract Interpretation (S), \emph{Guided Static Analysis} (G),
\emph{Path-focusing} (PF), our combined technique (G+PF), and its version using
disjunctive invariants (DIS). For
instance, \textbf{G/S} compares the benefits of \emph{Guided Static Analysis}
over the classic Abstract interpretation algorithm.
$\subsetneq$, $\supsetneq$ and ``unc.'' defined as in Fig.~\ref{fig:techniques}.}
\label{tab:techniques}
\end{table*}

>>>>

\JH{$99990)$, ca designe bien une borne d'intervalle ouvert ? J'ai plus
l'habitude a la notation $99990[$}
\MM{Je crois que les anglophones utilisent plutôt $99990)$, mais
  d'après Wikipedia ``Interval (mathematics)'', les deux sont acceptés
  par le standard ISO}
Widening iterations converge to $\lstinline|x_old| \in (-\infty,+\infty)$; let us now see why decreasing iterations fail to recover the desired invariant.
The \lstinline|x > x_old+10| test at line~6, if taken, yields $\lstinline|x_old|
\in (-\infty,99990)$; followed by \lstinline|x = x_old+10|, we obtain
$\lstinline|x| \in (-\infty,\allowbreak 100000)$, and the same after union with the no-operation ``else'' branch. Line~7 yields $\lstinline|x| \in (-\infty,+\infty)$.

@@

For a loop header $p_i$, this technique computes a disjunctive invariant
$\bigvee_{1\leq j \leq m} X_{i,j}$. We denote by $n$ the number of
distinct paths in the loop.
To perform the analysis, one chooses an integer $\delta \in [1,m]$, and
a mapping function $\sigma: [1,m] \times [1,n] \mapsto [1,m]$.
\MM{C'est moi ou le $\tau_{i,k}$ ne correspond pas exactement à la
  notation du début ? Au début, $\tau_{i,k}$ est la transition de $i$
  à $k$, alors que là, on est dans le multigraphe donc il y a
  plusieurs arrêtes entre deux points. Si je comprends bien,
  $\tau_{i,k}$ est le $k$-ième chemin partant de $i$, c'est ça ?
  Je mettrais bien ``... and for the $k$-th path starting from $i$,
  denoted $\tau_{i,k}$, the image...'' si c'est bien ça. D'ailleurs,
  faut-il forcément parler de boucle ici ? La technique ne s'applique
  pas à des chemins quelconques ?}
  For each
abstract value of the disjunctive invariant, and for each path $\tau_{i,k}$ in the loop, the
image of $X_{i,j}$ by the path is joined with
$X_{i,\sigma(j,k)}$.
Initially, the $\delta$-th abstract value contains the initial states of
$p_i$, and all other abstract values contain~$\emptyset$.

@@

Our improvement consists in a modification of the SMT formula we solve in
\ref{sec:guided_multigraph}.
We introduce in this formula Boolean variables $\{d_j, 1 \leq j \leq m\}$, so
that we can easily find in the model which abstract value of the
disjunction of the source point has
to be chosen to make the invariant of the destination grow.
The resulting formula that is given to the SMT solver is defined
by $g(p_i)$.
When the formula is satisfiable, we know that the index $j$ of the starting
disjunct that has to be chosen is the one for which the associate Boolean value
$d_j$ is \emph{true} in the model. Then, we can easily compute the value of 
$\sigma(j,k)$, thus know the index of the disjunct to join with.

@@

\MM{Je trouve ça assez frustrant d'arriver ici et de ne toujours pas
  savoir quel $\sigma$ est choisi. A mon avis, on peut remonter la
  description du $\sigma$ dynamique de la section suivante à ici. Ça
  évite de le présenter comme un petit détail d'implémentation alors
  que c'est quand même une vraie contrib par rapport au papier
  original.}

@@

\subsection{Analysis algorithm}
\label{sec:analysis-algorithm}
LLVM bitcode is in \emph{static single assignment} (SSA) form: a given scalar variable is given a value at a single syntactic point in the program. In concrete terms, an assignment \lstinline|x=2*x+1;| gets translated into a definition $x_2 = 2x_1+1$, with distinct variables $x_1$ and $x_2$ corresponding to the same original variable \lstinline|x| at different points in the program.
LLVM makes it easy to follow definition-use and use-definition chains: for a given variable (say, $x_2$) one can immediately obtain its definition (say, $2x_1+1$).
One may see conversion to SSA form as a static precomputation of some of the symbolic propagations proposed by \citet{DBLP:conf/vmcai/Mine06} to enhance the precision of analyses.

@@

\MM{cf. ci-dessus, j'aurais mis ça dans la section invariants
  disjonctifs}
For the computation of disjunctive invariants, we experimented with a heuristic
of dynamic construction of the $\sigma$ function, adapted from
\cite{DBLP:conf/pldi/GulwaniZ10}.  We start with one single
disjunct per control point ($m=1$), and we start with an empty map $\sigma$. We also
define a maximal bound on the number of disjunct per control point, noted $M$.
$\sigma(j,k)$ is then constructed on the fly only when needed.
When $\sigma(j,k)$ needs to be computed, we first compute the image of the
abstract value $X_k$ by the path indexed by $j$, and try to find an existing
disjunct of index $k'$ so that the least upper bound of the two abstract values is
exactly the union of them. If such an index exists, then we set $\sigma(j,k) = k'$.
Otherwise:
\begin{itemize}
	\item if $m < M$, we increase $m$ by $1$ and define $\sigma(j,k) = m$
	\item if $m = M$, we define $\sigma(j,k) = M$
\end{itemize}

@@

\begin{table*}
\begin{center}
\setlength{\tabcolsep}{0.75ex}
\begin{tabular}{|l
|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}%
|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}%
|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}%
|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}%
|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}%
|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}|} \hline
\multicolumn{1}{|c|}{\textbf{Benchmark}}
& \multicolumn{3}{c|}{\textbf{G/S}}
& \multicolumn{3}{c|}{\textbf{PF/S}}
& \multicolumn{3}{c|}{\textbf{PF/G}}
& \multicolumn{3}{c|}{\textbf{G+PF/PF}}
& \multicolumn{3}{c|}{\textbf{G+PF/G}}
& \multicolumn{3}{c|}{\textbf{DIS/G+PF}} \\ %\cline{2-16}
& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.} \\
 \hline
 \input{table1}
\end{tabular}
\end{center}
\caption{Result of the comparison of the various techniques described in this
paper: classic Abstract Interpretation (S), \emph{Guided Static Analysis} (G),
\emph{Path-focusing} (PF), our combined technique (G+PF), and its version using
disjunctive invariants (DIS). For
instance, \textbf{G/S} compares the benefits of \emph{Guided Static Analysis}
over the classic Abstract interpretation algorithm.
$\subsetneq$, $\supsetneq$ and ``unc.'' defined as in Fig.~\ref{fig:techniques}.}
\label{tab:techniques}
\end{table*}
